---
title: "The rminer package for regression"
author: Gabriele Venturato
output:
  pdf_document: default
bibliography: bibliography.bibtex
nocite: |
  @Cor10, @trevor2009elements,  @witten2016data
abstract: |
  The aim of this work is to have an insight into the *rminer* package for regression analysis. Starting from a brief theoretical introduction, towards the description of the main functions of the package, and concluding with a simple case study to show how the package can be used.
---


# Introduction
## Regression
Regression is the problem of learning a *functional relationship* between variables using a dataset where the specific functional form learned depends on the choice of the model (it can be linear or not). The parameters of the function are learned using the *explanatory variables (features)* into the training set, and then performance are evaluates testing the model on the test set. The aim of a regression model --- as opposed to a classification model --- is to perform a *numeric prediction* based on the features in input.

## Linear Regression


## Random Forest

# The rminer package
The goal of this package is to facilitate the use of data mining algorithms for classification and regression. It offers a short and coherent set of functions in order to easily develop a project, letting the user to follow in particular three CRISP-DM stages: *data preparation*, *modeling* and *evaluation*.

The package can be installed and loaded with:
```r
install.packages("rminer")
```

And loaded with:
```{r}
library(rminer)
```

As usual, a complete list of all functions available can be found in the documentation of the package:
```r
help(package=rminer)
```

For the purpose of this work instead of reporting what can be found easily --- and with more details --- inside the help, I preferred to report a brief list of the function organized by their purpose, in order to quickly move toward the practical example that is more useful to show the package capabilites.

First of all, for the data preparation phase, after having loaded the dataset, the first function that can be used are mainly:

* \texttt{delevels(x, levels, label = NULL)} -- reduce or replace factor *x* with *levels*, with an optional new label;
* \texttt{imputation(imethod = "value", D, Attribute = NULL, Missing = NA, Value = 1)} -- perform imputation to remove missing values from dataset *D* and from a specific attribute, with the value specified.
* \texttt{CaseSeries} -- create a data.frame from a time series (vector) using a sliding window. This function is not used in this work and its behavior can be further analized in official documentation.





# Case Study
```{r include=FALSE}
library(ggplot2)
library(kknn)
library(cowplot)
library(corrplot)
```
In this section I'll give a quick tour towards the main functionalities of rminer by mean of a dataset.

## The dataset
The dataset is about Life Expectancy and can be found in Kaggle [@LEdataset]. This dataset is available thanks to the World Health Organization who keeps track of the health status for all countries. It contains data about 193 countries from the year 2000 to 2015. All data column have a pretty self-explanatory name. For more details one can have a look into the official website from which the dataset has been taken.

For the purpose of this work a quick idea about the data can be achieved with the summary function in R, after lodaing it.
```{r}
lifeexp.df = read.csv("Life Expectancy Data.csv")
str(lifeexp.df)
summary(lifeexp.df) # here we can see NAs
```

From here can be seen that there are 22 columns and that some of them have missing values that will need to be taken care of. The purpose is to use the *Life.expectancy* variable as dependent, and all the others as predictors.

An important note here about the package is that since the country variable is stores as a factor, using this dataset I've find out that rminer can't handle factors with more than 53 levels, so I transformed the country factor as a numerical.



## The model

## Evaluation

# Conclusion

# References